First steps in synthetic algebraic geometry

### 2017-02-07 slides-warwick2017.tex
Slides finished

\frame[t]{\frametitle{Outline}\scriptsize\begin{itemize}\item[]\tableofcontents\end{itemize}}

@@

  \begin{changemargin}{2.5em}{2.5em}
    We describe how the internal language of certain toposes, the
    associated little and big Zariski toposes of a scheme, can be used
    to give simpler definitions and more conceptual proofs of the basic
    notions and observations in algebraic geometry.

    The starting point is that, from the "internal point of view" of the
    little Zariski topos, sheaves of rings and sheaves of modules look
    just like plain rings and plain modules. In this way, some concepts
    and statements of scheme theory can be reduced to concepts and
    statements of intuitionistic linear algebra. This simplifies working
    with sheaves and brings conceptual clarity.

    The internal language of the big Zariski topos goes even further. It
    incorporates Grothendieck's functor-of-points philosophy in order to
    cast modern algebraic geometry, relative to an arbitrary base
    scheme, in a naive language reminiscient of the classical Italian
    school. The base scheme looks like the one-element set from this
    point of view.

    The talk gives an introduction to this topos-theoretic point of view
    of algebraic geometry in the hope that it's useful to working
    algebraic geometers. No prior knowledge about toposes is supposed.
  \end{changemargin}
}

@@

\frame[t]{\frametitle{The internal language of~$\Sh(X)$}
  \small
  Let~$X$ be a topological space. Then we recursively define
  \[ U \models \varphi \quad \text{(``$\varphi$ holds on~$U$'')} \]
  for open subsets~$U \subseteq X$ and formulas~$\varphi$.
  % Write~``$\Sh(X) \models \varphi$'' to mean~$X \models \varphi$.
  \pause
  \[ \renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\ }c@{\ }l@{}}
    U \models f = g \? \F &\Ll& f|_U = g|_U \in \F(U) \\
    U \models \varphi \wedge \psi &\Ll&
      \text{$U \models \varphi$ and $U \models \psi$} \\
    U \models \varphi \vee \psi &\Ll&
      \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt} \\
    && \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
    && \quad\quad \text{$U_i \models \varphi$ or $U_i \models \psi$} \\
    U \models \varphi \Rightarrow \psi &\Ll&
      \text{for all open~$V \subseteq U$: } 
    \text{$V \models \varphi$ implies $V \models \psi$} \\
    U \models \forall f \? \F\_ \varphi(f) &\Ll&
      \text{for all sections~$f \in \F(V), V \subseteq U$: $V \models
      \varphi(f)$} \\
    U \models \exists f \? \F\_ \varphi(f) &\Ll&
      \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
    && \quad\quad \text{there exists~$f_i \in \F(U_i)$ s.\,th.
    $U_i \models \varphi(f_i)$}
  \end{array} \]
}

@@

\frame[t]{\frametitle{The internal language of~$\Sh(X)$}
  \begin{block}{Crucial property: Locality}
    If~$U = \bigcup_i U_i$, then
      $U \models \varphi$ iff $U_i \models
      \varphi$ for all~$i$.
  \end{block}

@@

  \pause\pause
  \begin{block}{A first glance at the constructive nature}
  \begin{itemize}
    \item $U \models f = 0$
      \tabto{3.5cm} iff $f|_U = 0 \in \Gamma(U, \F)$.
    \item $U \models f = 0 \vee g = 0$
      \tabto{3.5cm} iff on a cover~$U = \bigcup_i U_i$,\\
      \tabto{3.5cm} \qquad\qquad $f|_{U_i} = 0$ or~$g|_{U_i} = 0$.
    \item $U \models \neg\neg(f = 0)$
      \tabto{3.5cm} iff~$f = 0$ on a dense open subset of~$U$.
  \end{itemize}
  \end{block}
}

@@

\section[Little Zariski Topos]{The little Zariski topos of a scheme}

@@

  \begin{block}{Basic look and feel}
  \begin{itemize}
    \item Internally, the structure sheaf~$\O_X$ looks like \[ \text{
    an ordinary ring.} \]
    \item Internally, a sheaf of~$\O_X$-modules looks like \[
    \text{an ordinary module on that ring.} \]
  \end{itemize}
  \end{block}
}

@@

\begin{frame}\frametitle{The objective}
  \slogan{\justifying Understand notions and statements of \hil{algebraic geometry} as
  notions and statements of (intuitionistic) \hil{commutative
  algebra} internal to the \hil{little Zariski topos}.}

@@

  Further topics regarding the little Zariski topos:
  \begin{itemize}
    \item Transfer principles~$M \leftrightarrow M^\sim$
    \item The curious role of affine open subsets
    \item Quasicoherence
    \item Spreading from points to neighbourhoods
    \item The relative spectrum
  \end{itemize}
\end{frame}

@@

  \centering
  \hil{Answer:} $M^\sim$ inherits those properties of $M$ which are \\
  \hil{stable under localization}.
  \par
\end{frame}

@@

  Prime filters are important in constructive mathematics because localizing
  them gives rise to local rings. In contrast, localizing a ring at the
  complement of a prime ideal doesn't usually result in a local ring.

  To construct the universal localization of~$A$, one doesn't pick a particular
  prime filter~$F$ to construct~$A[F^{-1}]$. Instead, one picks the
  \emph{generic prime filter}~$\F$. This filter doesn't live in~$\Set$, but
  in~$\Sh(\Spec A)$.
  \par
}

\begin{frame}\frametitle{The curious role of affine open subsets}
  \hil{Question:} Why do the following identities hold, for quasicoherent
  sheaves~$\E$ and~$\F$ and affine open subsets~$U$?
  \begin{align*}
    (\E/\F)(U) &= \E(U) / \F(U) \\
    (\E \otimes_{\O_X} \F)(U) &= \E(U) \otimes_{\O_X(U)} \F(U) \\
    \E_\mathrm{tors}(U) &= \E(U)_\mathrm{tors} \quad\text{(sometimes)} \\
    \K_X(U) &= \operatorname{Quot} \O_X(U) \quad\text{(sometimes)}
  \end{align*}
  \pause
  \hil{A calculation:}
  \begin{align*}
    M^\sim \otimes_{\O_U} N^\sim &=
    \ull{M}[\F^{-1}] \otimes_{\ull{A}[\F^{-1}]} \ull{N}[\F^{-1}] =
    (\ull{M} \otimes_\ull{A} \ull{N})[\F^{-1}] \\
    &= (\ull{M \otimes_A N})[\F^{-1}] =
    (M \otimes_A N)^\sim.
  \end{align*}
  \pause
  \hil{Answer:} Because localization commutes with quotients, tensor products,
  torsion submodules (sometimes), \ldots
\end{frame}

\subsection[Quasicoherence]{Quasicoherence of sheaves of modules}

@@

\begin{frame}\frametitle{Generic freeness}
  \vspace*{-1.6em}
  \[ \xymatrixcolsep{3pc}\xymatrixrowsep{3pc}\xymatrix{
    & \ar@{-}[d]^{\substack{\text{finitely}\\\text{generated}}} M \\
    R \ar[r]_{\text{of finite type}} & S
  } \]
  If $R$ is reduced ($x^n = 0 \Rightarrow x = 0$), there is a \hil{dense open}~$U \subseteq \Spec R$ such that for any $f \in R$ with~$D(f)
  \subseteq U$,
  \begin{itemize}
    \item $S[f^{-1}]$ and $M[f^{-1}]$ are free modules over $R[f^{-1}]$,
    \item $R[f^{-1}] \to S[f^{-1}]$ is of finite presentation, and
    \item $M[f^{-1}]$ is finitely presented as a module over~$S[f^{-1}]$.
  \end{itemize}
  \pause

@@

@@

  \medskip
  Form limits in the category of \hil{locally ringed locales}
  by \hil{relocalizing} the corresponding limit in ringed locales.
\end{frame}

@@

\section[Big Zariski topos]{The big Zariski topos of a scheme}

\begin{frame}\frametitle{The big Zariski topos}
  \begin{block}{Definition}The \hil{big Zariski topos} $\Zar(S)$ of a scheme~$S$ is the
  category $\Sh(\Aff/S)$. It consists of functors
  $(\Aff/S)^\op \to \Set$ satisfying the sheaf condition that
  \[ F(U) \to
    \prod_i F(U_i) \rightrightarrows
    \prod_{j,k} F(U_j \cap U_k) \]
  \\[-1.5em]
  is a limit diagram for any affine scheme~$U = \bigcup_i U_i$ over~$S$.
  \end{block}

  \begin{itemize}
    \item For an~$S$-scheme~$X$, its functor of points $\ull{X} =
    \Hom_S(\cdot,X)$ is an object of~$\Zar(S)$. It feels like \hil{the set of
    points} of~$X$.
    \item Internally, $\affl$ (given by $\affl(X) = \O_X(X)$)
    looks like a field:
    \[ \Zar(S) \models \forall x\?\affl\_ x \neq 0 \Longrightarrow \speak{$x$ invertible} \]
  \end{itemize}

\subsection{Some internal constructions}

\begin{frame}\frametitle{Some internal constructions}
  \begin{itemize}
  \item $\mathbb{P}^n_S = \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\,
    x_0 \neq 0 \vee \cdots \vee x_n \neq 0 \}/(\affl)^\times.$

  \item \ \\[-1.2em]\mbox{$\RelSpec_S R = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
  \text{set of $\affl$-rational points of $R$}.$}

  \item $TX = \Hom(\Delta, X)$,
  where~$\Delta = \{ \varepsilon : \affl \,|\, \varepsilon^2 = 0 \}$.
  \pause

  \item In particular, any map $\affl \to \affl$ is given by a polynomial.
  \pause

  \item A subset $U \subseteq X$ is open if and only if for any $x : X$
  there exists an ideal $J \subseteq \affl$ such that $\affl/J$ is quasicoherent and
  \[ x \in U \quad\Longleftrightarrow\quad J = (1). \]

  \item A subset $U \subseteq X$ is qc-open if and only if for any $x : X$
  there exist $f_1,\ldots,f_n \subseteq \affl$ such that
  \[ x \in U \quad\Longleftrightarrow\quad \exists i\_ \text{$f_i$ is invertible}. \]
  \end{itemize}
\end{frame}

% Add two bigger case studies:
% * Construction of the Grassmannian
% * f : Delta --> X such that f(0) in U implies that f factors over U
%
% And add the following smaller things:
% * qc, sep, qs
% * universally closed
% * surjective

@@

\begin{frame}\frametitle{Semi-open and open tasks}
  \begin{itemize}
  \item Do cohomology in the little Zariski topos; exploit that higher direct
  images look like ordinary sheaf cohomology from the internal point of view.
  \item Do cohomology in the big Zariski topos.
  \item Understand more subtoposes of the big Zariski topos.
  \item Derive suitable axioms for synthetic algebraic geometry.
  \end{itemize}

  {\vspace{-0.1em}\centering
  \rotatebox{90}{\tiny\scalebox{0.5}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
  \includegraphics[scale=0.20]{images/external-internal-small}
  \par\medskip\vspace{-0.1em}}
\end{frame}

\begin{frame}
  \slogan{\hil{Understand notions and statements of algebraic geometry
  as notions and statements of algebra internal to appropriate toposes.}}

  {\vspace{-0.1em}\centering
  \rotatebox{90}{\tiny\scalebox{0.5}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
  \includegraphics[scale=0.20]{images/external-internal-small}
  \par\medskip\vspace{-0.1em}}

  \begin{itemize}
    \item Simplify proofs and gain conceptual understanding.
    \item Understand relative geometry as absolute geometry.
    \item Develop a synthetic account of scheme theory.
    \item Contribute to constructive algebra.
  \end{itemize}

>>>>

\frame[t]{\frametitle{Outline}\small
  \begin{columns}[t]
    \begin{column}{0.6\textwidth}
      \vspace*{-1.5em}
      \begin{itemize}\item[]\tableofcontents\end{itemize}
      \par
    \end{column}
    \begin{column}{0.4\textwidth}
      \centering
      \begin{minipage}{0.70\textwidth}
        \begin{exampleblock}{}
          \justifying
          Any finitely generated vector space does \emph{not not} possess a basis.
        \end{exampleblock}
      \end{minipage}
      \medskip

      \begin{minipage}{0.70\textwidth}
        \begin{exampleblock}{}
          \justifying
          Any sheaf of modules of finite type on a reduced scheme is locally free
          \emph{on a dense open subset}.
        \end{exampleblock}
        \centering
        \tiny Ravi Vakil: ``Important hard exercise'' (13.7.K).
        \par
      \end{minipage}
    \end{column}
  \end{columns}
}

@@

  \begin{changemargin}{2.5em}{2.5em}
    (new abstract)
  \end{changemargin}
}

@@

\frame[t]{\frametitle{The internal language of~$\boldsymbol{\Sh(X)}$}
  \small
  Let~$X$ be a topological space. Then we recursively define
  \[ U \models \varphi \quad \text{(``$\varphi$ holds on~$U$'')} \]
  for open subsets~$U \subseteq X$ and formulas~$\varphi$.
  % Write~``$\Sh(X) \models \varphi$'' to mean~$X \models \varphi$.
  \pause
  \[ \renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\ }c@{\ }l@{}}
    U \models f = g \? \F &\Ll& f|_U = g|_U \in \F(U) \\
    U \models \varphi \wedge \psi &\Ll&
      \text{$U \models \varphi$ and $U \models \psi$} \\
    U \models \varphi \vee \psi &\Ll&
      \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt} \\
    && \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
    && \quad\quad \text{$U_i \models \varphi$ or $U_i \models \psi$} \\
    U \models \varphi \Rightarrow \psi &\Ll&
      \text{for all open~$V \subseteq U$: } 
    \text{$V \models \varphi$ implies $V \models \psi$} \\
    U \models \forall f \? \F\_ \varphi(f) &\Ll&
      \text{for all sections~$f \in \F(V), V \subseteq U$: $V \models
      \varphi(f)$} \\
    U \models \exists f \? \F\_ \varphi(f) &\Ll&
      \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
    && \quad\quad \text{there exists~$f_i \in \F(U_i)$ s.\,th.
    $U_i \models \varphi(f_i)$}
  \end{array} \]
}

@@

\frame[t]{\frametitle{The internal language of~$\boldsymbol{\Sh(X)}$}
  \begin{block}{Crucial property: Locality}
    If~$U = \bigcup_i U_i$, then
      $U \models \varphi$ iff $U_i \models
      \varphi$ for all~$i$.
  \end{block}

@@

  \pause\pause
  \begin{block}{A first glance at the constructive nature}
  \begin{itemize}
    \item $U \models f = 0$
      \tabto{2.9cm} iff $f|_U = 0 \in \Gamma(U, \F)$.
    \item $U \models \neg\neg(f = 0)$
      \tabto{2.9cm} iff~$f = 0$ on a dense open subset of~$U$.
  \end{itemize}
  \end{block}
}

@@

\section[Little Zariski]{The little Zariski topos of a scheme}

@@

  \begin{itemize}
    \item Internally, the structure sheaf~$\O_X$ looks like \[ \text{
    an ordinary ring.} \]
    \item Internally, a sheaf of~$\O_X$-modules looks like \[
    \text{an ordinary module on that ring.} \]
  \end{itemize}
}

@@

\begin{frame}\frametitle{The objective}
  \slogan{\justifying Understand notions and statements of \hil{algebraic geometry} as
  notions and statements of (constructive) \hil{commutative
  algebra} internal to the \hil{little Zariski topos}.}

@@

  Further topics regarding the little Zariski topos:
  \begin{itemize}
    \item Transfer principles~$M \leftrightarrow M^\sim$
    \item Understanding generic freeness
    \item The curious role of affine open subsets
    \item Quasicoherence
    \item Spreading from points to neighbourhoods
    \item The relative spectrum
  \end{itemize}
\end{frame}

@@

  \centering
  \hil{Answer:} $M^\sim$ inherits those properties of $M$ which are \\
  \hil{stable under localisation}.
  \par
\end{frame}

@@

  Prime filters are important in constructive mathematics because localising
  them gives rise to local rings. In contrast, localising a ring at the
  complement of a prime ideal doesn't usually result in a local ring.

  To construct the universal localisation of~$A$, one doesn't pick a particular
  prime filter~$F$ to construct~$A[F^{-1}]$. Instead, one picks the
  \emph{generic prime filter}~$\F$. This filter doesn't live in~$\Set$, but
  in~$\Sh(\Spec A)$.
  \par
}

\subsection[Generic freeness]{Generic freeness}

@@

\begin{frame}\frametitle{Generic freeness}
  \vspace*{-1.6em}
  \[ \xymatrixcolsep{3pc}\xymatrixrowsep{3pc}\xymatrix{
    & \ar@{-}[d]^{\substack{\text{finitely}\\\text{generated}}} M \\
    R \ar[r]_{\text{of finite type}} & S
  } \]
  If $R$ is reduced ($x^n = 0 \Rightarrow x = 0$), there is a \hil{dense open subset} $U \subseteq \Spec R$ such that for any $f \in R$ with~$D(f)
  \subseteq U$,
  \begin{enumerate}
    \item $S[f^{-1}]$ and $M[f^{-1}]$ are free modules over $R[f^{-1}]$,
    \item $R[f^{-1}] \to S[f^{-1}]$ is of finite presentation, and
    \item $M[f^{-1}]$ is finitely presented as a module over~$S[f^{-1}]$.
  \end{enumerate}
  \pause

@@

\section[Big Zariski]{The big Zariski topos of a scheme}

\begin{frame}\frametitle{The big Zariski topos}
  \begin{block}{Definition}The \hil{big Zariski topos} $\Zar(S)$ of a scheme~$S$ is the
  category $\Sh(\Aff/S)$. It consists of functors
  $(\Aff/S)^\op \to \Set$ satisfying the gluing condition that
  \[ F(U) \to
    \prod_i F(U_i) \rightrightarrows
    \prod_{j,k} F(U_j \cap U_k) \]
  \\[-0.6em]
  is a limit diagram for any affine scheme~$U = \bigcup_i U_i$ over~$S$.
  \end{block}

  \begin{itemize}
    \item For an~$S$-scheme~$X$, its functor of points $\ull{X} =
    \Hom_S(\cdot,X)$ is an object of~$\Zar(S)$. It feels like \hil{the set of
    points} of~$X$.
    \item Internally, $\affl$ (given by $\affl(X) = \O_X(X)$)
    looks like a field:
    \[ \Zar(S) \models \forall x\?\affl\_ x \neq 0 \Longrightarrow \text{$x$ invertible} \]
  \end{itemize}

\subsection{Synthetic constructions}

\begin{frame}\frametitle{Synthetic constructions}
  \begin{itemize}
  \item $\mathbb{P}^n_S = \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\,
    x_0 \neq 0 \vee \cdots \vee x_n \neq 0 \}/(\affl)^\times.$ \\[1em]

  \item \ \\[-1.2em]\mbox{$\RelSpec R = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
  \text{set of $\affl$-rational points of $R$}.$} \\[1em]

  \item $TX = \Hom(\Delta, X)$,
  where~$\Delta = \{ \varepsilon : \affl \,|\, \varepsilon^2 = 0 \}$.
  \end{itemize}
\end{frame}

\subsection{Synthetic formulation of properties}

\begin{frame}{Synthetic formulation of properties}
  \begin{itemize}
  \item \ \\[-1.2em]\mbox{$\RelSpec R = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
  \text{set of $\affl$-rational points of $R$}.$}

  \item In particular, any map $\affl \to \affl$ is given by a polynomial. \bigskip
  \pause

  \item A subset $U \subseteq X$ is qc-open if and only if for any $x : X$
  there exist $f_1,\ldots,f_n \in \affl$ such that
  $x \in U \Longleftrightarrow \exists i\_ f_i \neq 0$.

  \item Open subsets are $\neg\neg$-stable:
  $\neg\neg(x \in U) \Longrightarrow x \in U$.

  \item If $\gamma : \Delta \to X$ is tangent vector with $\gamma(0) \in U$,
  then~$\gamma(\varepsilon) \in U$ for all~$\varepsilon \in \Delta$.
  \bigskip\pause

  \item $X$ is separated if and only if for any $x,y : X$,
  there exists a quasicoherent ideal~$J \subseteq \affl$ such that
  $x = y \Longleftrightarrow J = (0)$.
  \end{itemize}
\end{frame}

% Add two bigger case studies:
% * Construction of the Grassmannian
%
% And add the following smaller things:
% * qc, sep, qs
% * universally closed
% * surjective

\begin{frame}\frametitle{Semi-open and open tasks}
  \begin{itemize}
  \item Do cohomology in the little Zariski topos; exploit that higher direct
  images look like ordinary sheaf cohomology from the internal point of view.
  \item Do cohomology in the big Zariski topos.
  \item Understand subtoposes of the big Zariski topos.
  \end{itemize}

  {\vspace{-0.1em}\centering
  \rotatebox{90}{\tiny\scalebox{0.5}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
  \includegraphics[scale=0.28]{images/external-internal-small}
  \par\medskip\vspace{-0.1em}}
\end{frame}

\begin{frame}{}

  {\vspace{-0.1em}\centering
  \rotatebox{90}{\tiny\scalebox{0.5}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
  \includegraphics[scale=0.17]{images/external-internal-small}
  \par\medskip\vspace{-0.1em}}

  \slogan{\hil{Understand notions and statements of algebraic geometry
  as notions and statements of algebra internal to the little Zariski topos.}}

  \slogan{\hil{Develop a synthetic account of algebraic geometry.}}

  \begin{itemize}
    \item Simplify proofs and gain conceptual understanding.
    \item Understand relative geometry as absolute geometry.
    \item Contribute to constructive algebra.
  \end{itemize}

\begin{frame}\frametitle{The curious role of affine open subsets}
  \hil{Question:} Why do the following identities hold, for quasicoherent
  sheaves~$\E$ and~$\F$ and affine open subsets~$U$?
  \begin{align*}
    (\E/\F)(U) &= \E(U) / \F(U) \\
    (\E \otimes_{\O_X} \F)(U) &= \E(U) \otimes_{\O_X(U)} \F(U) \\
    \E_\mathrm{tors}(U) &= \E(U)_\mathrm{tors} \quad\text{(sometimes)} \\
    \K_X(U) &= \operatorname{Quot} \O_X(U) \quad\text{(sometimes)}
  \end{align*}
  \pause
  \hil{A calculation:}
  \begin{align*}
    M^\sim \otimes_{\O_U} N^\sim &=
    \ull{M}[\F^{-1}] \otimes_{\ull{A}[\F^{-1}]} \ull{N}[\F^{-1}] =
    (\ull{M} \otimes_\ull{A} \ull{N})[\F^{-1}] \\
    &= (\ull{M \otimes_A N})[\F^{-1}] =
    (M \otimes_A N)^\sim.
  \end{align*}
  \pause
  \hil{Answer:} Because localisation commutes with quotients, tensor products,
  torsion submodules (sometimes), \ldots
\end{frame}

\subsection{Quasicoherence}

@@

  \medskip
  Form limits in the category of \hil{locally ringed locales}
  by \hil{relocalising} the corresponding limit in ringed locales.
\end{frame}

@@

@@
